---
title: "nCounter_RNA_Analysis"
author: "Ned Cauley"
date: "4/16/24"
output: html_document
---

# This markdown and analysis based on https://github.com/bhattacharya-a-bt/CBCS_normalization/blob/master/sabry_analysis.R
# The above analysis is from the publication https://academic.oup.com/bib/article/22/3/bbaa163/5891144?login=true

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up libraries and QC functions libraries

```{r Libraries and RUVseq functions, message=FALSE, warning=FALSE}
library(NanoStringQCPro)
library(ggplot2)
library(EnvStats)
library(MASS)
library(RUVSeq)
library(DESeq2)
library(limma)
library(matrixStats)
library(limma)
library(NanoStringNorm)
library(readxl)
library(plotly)
library(pheatmap)
library(ggrepel)


# The library in RSW to use is located at
# /home/cauleyes/ccbr-data/renv_cache/ncounter_rna/Snapshot-environment_method/renv/library/R-4.1/x86_64-pc-linux-gnu

# Set the working directory to your project folder if it is not set already
setwd("/home/cauleyes/ccbr-data/users/Ned/nCounter_rna/CCBR1245_Ramaswami_nCounter")

# Functions for RUVseq analysis
#source("R/nanostring_RUV_functions.R")
source("R/ncounter_analysis_functions.R")

# Load the annotation file
#annotation_file_path <- "ncounter_rna_annotation_edit_correction_RR.xlsx"
annotation_file_path <- "4_15_24_ncounter_rna_annotation_edit_NC.xlsx"
annotation_df <- read_excel(annotation_file_path)

```

```{r Define the contrast for the DEG analysis, message=FALSE, warning=FALSE}

# Define the contrast and levels from the annotation columns for QC and DEG analysis
contrasts.list <- list(
  "CD4_count" = c("High", "Low"), 
  "KS_alone" = c("Y", "N"), 
  "KSHV_VL" = c("1", "0"), 
  "HIV_VL" = c("Y", "N"), 
  "response" = c("Y", "N")
)

#contrast <- "CD4_group"
#contrast <- "KS_alone"
#contrast <- "KSHV_VL_Bin"


# Levels are: Reference, Effect. 
# This means a positive log2foldchnage = lower reference, higher effect
#contrast_levels <- c("High", "Low")
#contrast_levels <- c("Y", "N")
#contrast_levels <- c("0", "1")





```


## Setup the initial dataframes with read counts and metadata/QC

```{r Initial Dataframes and QC,  message=FALSE, warning=FALSE}

rcc.files.path <- "/rstudio-files/ccbr-data/users/Ned/nCounter_rna/data/CCBR1245/rcc"

# Gather the names of the RCC files
#files.RCC = list.files("/rstudio-files/ccbr-data/users/Ned/nCounter_rna/data/CCBR1245/rcc")

# Process the RCC files as count, metadata, and feature data DFs
rcc.dfs <- process_RCC_files(rcc.files.path = rcc.files.path, 
                  annotation.df = annotation_df)

# Print a sumnmary of the DFs
head(rcc.dfs$raw.counts)
head(rcc.dfs$pData)
head(rcc.dfs$fData)

```

## Verify that housekeeping genes are not DEGs

```{r, message=FALSE, warning=FALSE}
###
#### CHECK IF HK ARE ASSOCIATED WITH PRIMARY PHENO
###

# Gather the read counts for housekeeping genes
hk_raw = rcc.dfs$raw.counts[rcc.dfs$hk.gene.list,]
pval = vector(length = nrow(hk_raw))

# Run general linear model to get pvalue for housekeeping
for (i in 1:nrow(hk_raw)){
  
  reg = glm.nb(as.numeric(hk_raw[i,]) ~ as.factor(rcc.dfs$pData[[contrast]]))
  pval[i] = coef(summary(reg))[2,4]
  
}

print("Housekeeping genes:")
print(rownames(hk_raw))

# Number of housekeeping that are DE
print("Number of housekeeping genes that are DE:")
sum(pval <= .05)

```

## Run RUVseq differential expression analysis

```{r, message=FALSE, warning=FALSE}

# Create a new folder for the run date
run.folder <- "4_16_2024/"

for(contrast in names(contrasts.list)){
  
  contrast.levels <- contrasts.list[[contrast]]
  
  # Run DESEQ with RUVseq normalization for k sources of variability
  deseq.output <- run_RUV_DESEQ2(contrast = contrast, 
                 contrast.levels = contrast.levels, 
                 annotation.data = rcc.dfs$pData, 
                 raw.counts = rcc.dfs$raw.counts, 
                 feature.data = rcc.dfs$fData, 
                 k = 1, 
                 additional.covar = NULL, 
                 hk.genes = rcc.dfs$hk.gene.list, 
                 exclude.hk.genes = NULL)
  
  
  ## Make a volcano plot for the DEGs
  
  # Remove the control probes
  deg.results.df <- deseq.output$all.de.results[!grepl("^POS_|^NEG_", 
                                                       deseq.output$all.de.results), ]
  
  
  # Create a column for direction of DEGs
  deg.results.df$de_direction <- "NONE"
  deg.results.df$de_direction[deg.results.df$padj < 0.05 & 
                               deg.results.df$log2FoldChange > 1] <- "UP"
  deg.results.df$de_direction[deg.results.df$padj < 0.05 & 
                               deg.results.df$log2FoldChange < -1] <- "DOWN"
  
  deg.results.df$deglabel <- ifelse(deg.results.df$de_direction == "NONE", NA, deg.results.df$gene)
  
  # Compute the scale for the volcano x-axis
  log2.scale <- max(abs(deg.results.df$log2FoldChange))
  
  # Make the volcano plot
  
  volcano.plot <- ggplot(data = deg.results.df, aes(x = log2FoldChange, 
                                                    y = -log10(padj), 
                                                    col = de_direction, 
                                                    label = deglabel)) +
    geom_vline(xintercept = c(-1, 1), col = "gray", linetype = 'dashed') +
    geom_hline(yintercept = -log10(0.05), col = "gray", linetype = 'dashed') + 
    xlim(-7.5, 7.5) + 
    labs(x = paste(contrast, contrast.levels[[1]], "versus", contrast.levels[[2]]),
         y = "-log10 adjusted p-value", 
         title = paste0(contrast, " DE Analysis")) + 
    geom_point(size = 2) +
    scale_color_manual(values = c("steelblue4", "grey", "violetred4"), 
                       labels = c(paste("Downregulated in", contrast_levels[[1]]), "Not significant",   paste("Upregulated in", contrast_levels[[1]]))) + 
    geom_text_repel(max.overlaps = Inf) + 
    xlim(-log2.scale-1, log2.scale+1) + 
    theme(plot.title = element_text(hjust = 0.5))
  
  
  export.volcano <- TRUE
  if(export.volcano == TRUE){
      ggsave(paste0("Results/", run.folder, "volcano_", contrast, "_adjpval.png"),
           plot = volcano.plot, 
           width = 8, 
           height = 8, 
           units = "in")
  }
  
  # Export Normalized read counts
  export_norm_counts <- TRUE
  if(export_norm_counts == TRUE){
    # Grab the normalized_counts
    normalized_counts <- as.data.frame(counts(deseq.output$dds, normalized = TRUE))
    
    # Add a gene column
    gene <- rownames(normalized_counts)
    normalized_counts <- cbind(gene, normalized_counts) 
    
    # Export
    write.csv(normalized_counts, file = paste0("Results/", run.folder, "RUV_normalized_counts.csv"), row.names = FALSE)
    
    
  }
  
  # Export the DE results
  export_sig_DEG_list <- TRUE
  if(export_sig_DEG_list == TRUE){
    write.csv(deseq.output$deg.list, file = paste0("Results/", run.folder, "DEG_list_",contrast,".csv"), row.names =   FALSE)
  }
 
  
  
  
  
   
}



```

# Unsupervised analysis

```{r}

# Plot PCA to check for batch effects

# Prepare the counts
rlog_dds <- rlog(deseq.output$dds)

# Annotation to view PCA by
PCA.group <- "Date"

# PCA Plot
PCA.plot <- plotPCA(rlog_dds, intgroup = PCA.group, ntop = 700)

# Plotly PCA for interactivity
PCA.plot.data <- plotPCA(rlog_dds, intgroup = PCA.group, ntop = 700, returnData = TRUE)
PCA.plotly <- plot_ly(data = PCA.plot.data, x = ~PC1, y = ~PC2, text = ~name, mode = "markers", color = ~factor(group)) %>%
  layout(title = "PCA Plot",
         xaxis = list(title = "PC1"),
         yaxis = list(title = "PC2"))

print(PCA.plotly)

# Export the PCA plot

export.PCA <- FALSE
if(export.PCA == TRUE){
  ggsave(paste0("Results/QC/PCA.pdf"), plot = PCA.plot)
}

```
# Get the read count information for CXCL8

```{r}

normalized_CXCL8 <- set@assayData$counts[rownames(set@assayData$counts) == "CXCL8", ]


CXCL8_df <- data.frame(CXCL8_normalized = normalized_CXCL8, SampleID = names(normalized_CXCL8))
rownames(CXCL8_df) = NULL

CXCL8_anno_columns <- c("SampleID","KS_alone","CD4_group")
subset_anno_df <- annotation_df[CXCL8_anno_columns]


CXCL8_df <- merge(CXCL8_df, subset_anno_df, by = "SampleID")

CXCL8_mean_KS_Y <- mean(CXCL8_df$CXCL8_normalized[CXCL8_df$KS_alone == "Y"], na.rm = TRUE)
CXCL8_mean_KS_N <- mean(CXCL8_df$CXCL8_normalized[CXCL8_df$KS_alone == "N"], na.rm = TRUE)

CXCL8_df$KS_alone_mean <- ifelse(CXCL8_df$KS_alone == "Y", CXCL8_mean_KS_Y, CXCL8_mean_KS_N)

CXCL8_mean_CD4_High <- mean(CXCL8_df$CXCL8_normalized[CXCL8_df$CD4_group == "High"], na.rm = TRUE)
CXCL8_mean_CD4_Low <- mean(CXCL8_df$CXCL8_normalized[CXCL8_df$CD4_group == "Low"], na.rm = TRUE)

CXCL8_df$CD4_group_mean <- ifelse(CXCL8_df$CD4_group == "High", CXCL8_mean_CD4_High, CXCL8_mean_CD4_Low)



CXCL8_deg <- ruv_deg_list_export[ruv_deg_list_export$gene == "CXCL8", ]

export_CXCL8_summary <- FALSE
if(export_CXCL8_summary == TRUE){
  write.csv(CXCL8_df, file = "Results/CXCL8_summary.csv", row.names = FALSE)
}


```

# Heatmap of Variable Genes

```{r}

# Prepare the normalized counts
log.norm.counts <- log2(counts(dds, normalized = TRUE) + 1)

# Prepare the annotation columns
anno.columns <- c("KS_alone", "CD4_group")
annotation_df <- as.data.frame(annotation_df)

# Ensure the row names are the sample names
rownames(annotation_df) <- annotation_df$SampleID
heatmap.annotation <- annotation_df[ ,anno.columns]

# Order the annotation
heatmap.annotation <- heatmap.annotation %>% 
  arrange(CD4_group) %>% arrange(KS_alone)

# Define the colors for the annotations
anno.colors = list(
  KS_alone = c(Y = "seagreen", N = "red4"), 
  CD4_group = c(High = "purple", Low = "coral")
  )

# Determine the most variable genes using the coefficient of variation
.calcCv <- function(x) {
  sd(x) / mean(x)
}
cv.vector <- apply(log.norm.counts, 1, .calcCv)

# Identify the NEG and POS probes
NEG.indices <- grep("NEG_", names(cv.vector))
POS.indices <- grep("POS_", names(cv.vector))
control.indices <- c(NEG.indices, POS.indices)

# Print the names of the control probes to be removed
control.probes <- names(cv.vector)[control.indices]
print(control.probes)

# Remove the control probes
cv.vector <- cv.vector[-control.indices]

# Sort the genes based on variability
cv.vector <- sort(cv.vector, decreasing = TRUE)

# Top variable genes
ngenes <- 50
goi <- names(cv.vector)[1:ngenes]

# Reorder the data columns to they match the annotation row order
row.order <- rownames(heatmap.annotation)
log.norm.counts <- log.norm.counts[, row.order]


# Top 50 genes
heatmap.top50 <- pheatmap(log.norm.counts[goi, ], 
         main = "Top 50 Variable Genes", 
         show_rownames = TRUE, 
         scale = "row", 
         show_colnames = TRUE,
         border_color = NA, 
         cluster_rows = TRUE, 
         cluster_cols = FALSE, 
         clustering_method = "average", 
         clustering_distance_rows = "correlation", 
         clustering_distance_cols = "correlation", 
         color = colorRampPalette(c("blue", "white", "red"))(120), 
         annotation_row = NA, 
         annotation_col = heatmap.annotation, 
         annotation_colors = anno.colors
         )

export.top50 <- FALSE
if(export.top50 == TRUE){
  
  ggsave("Results/heatmap_top50.png",
         plot = heatmap.top50, 
         width = 10, 
         height = 10, 
         units = "in")
  
}

# Create a heatmap with all genes
heatmap.all <- pheatmap(log.norm.counts, 
         main = "All Genes", 
         show_rownames = FALSE, 
         scale = "row", 
         show_colnames = TRUE,
         border_color = NA, 
         cluster_rows = TRUE, 
         cluster_cols = FALSE, 
         clustering_method = "average", 
         clustering_distance_rows = "correlation", 
         clustering_distance_cols = "correlation", 
         color = colorRampPalette(c("blue", "white", "red"))(120), 
         annotation_row = NA, 
         annotation_col = heatmap.annotation, 
         annotation_colors = anno.colors
         )

export.heatmap.all <- FALSE
if(export.heatmap.all == TRUE){
  
  ggsave("Results/heatmap_all.png",
         plot = heatmap.all, 
         width = 10, 
         height = 10, 
         units = "in")
  
}


# Create a heatmap for the DEGs from the DEG contrast

# Get a list of the DEGS
contrast_deg_list <- rownames(contrast_ruv_deg_list[contrast_ruv_deg_list$padj < 0.05, ])

# Loosen the cutoff if no DEGs found
if(length(rownames(contrast_deg_list)) == 0){ 
  
  print("No DEGs found for padj < 0.05, trying padj < 0.1")
  
  contrast_deg_list <- rownames(contrast_ruv_deg_list[contrast_ruv_deg_list$padj < 0.1, ])
}

if(length(rownames(contrast_deg_list)) == 0){ 
  
  print("No DEGs found for padj < 0.05, trying pvalue < 0.05")
  
  contrast_deg_list <- rownames(contrast_ruv_deg_list[contrast_ruv_deg_list$pvalue < 0.05, ])
  
}

# Remove the positive and negative probes
# Identify the NEG and POS probes
NEG.indices <- grep("NEG_", contrast_deg_list)
POS.indices <- grep("POS_", contrast_deg_list)
control.indices <- c(NEG.indices, POS.indices)

# Remove the control probes
if(length(control.indices) > 0){
  goi <- contrast_deg_list[-control.indices]
} else{
  goi <- contrast_deg_list
}


# Gather the annotation for the contrast
anno.columns <- c(contrast)
annotation_df <- as.data.frame(annotation_df)

# Ensure the row names are the sample names
rownames(annotation_df) <- annotation_df$SampleID
heatmap.annotation <- annotation_df[ ,anno.columns, drop = FALSE]

# Order the annotation
heatmap.annotation <- heatmap.annotation %>% arrange(.data[[contrast]])

# Reorder the data columns to they match the annotation row order
row.order <- rownames(heatmap.annotation)
log.norm.counts <- log.norm.counts[, row.order]

# Define the annotation colors for the contrast
anno.colors1 = list(
  KS_alone = c(Y = "seagreen", N = "red4")
  )

anno.colors2 = list(
  CD4_group = c(High = "purple", Low = "coral")
  )

heatmap.contrast <- pheatmap(log.norm.counts[goi, ], 
         main = paste0("DEGs for ", contrast), 
         show_rownames = TRUE, 
         scale = "row", 
         show_colnames = TRUE,
         border_color = NA, 
         cluster_rows = TRUE, 
         cluster_cols = FALSE, 
         clustering_method = "average", 
         clustering_distance_rows = "correlation", 
         clustering_distance_cols = "correlation", 
         color = colorRampPalette(c("blue", "white", "red"))(120), 
         annotation_row = NA, 
         annotation_col = heatmap.annotation, 
         annotation_colors = anno.colors2
         )


export.heatmap.contrast <- TRUE
if(export.heatmap.contrast == TRUE){
  
  ggsave(paste0("Results/heatmap_", contrast, ".png"),
         plot = heatmap.contrast, 
         width = 12, 
         height = 12, 
         units = "in")
  
}

print(heatmap.top50)
print(heatmap.all)
print(heatmap.contrast)



```


# Clustering of Viral Gene Probes

```{r}

# See reference vignette: https://bioconductor.org/packages/release/bioc/vignettes/PCAtools/inst/doc/PCAtools.html#introduction

library(stringr)
library(dplyr)
library(ggfortify)
library(PCAtools)

# Transform DESEQ normalized counts to variance stabilized
vst.dds <- varianceStabilizingTransformation(deseq.output$dds)

# Gather normalized counts as a df
normalized.counts <- as.data.frame(assay(vst.dds))

# Identify the KSHV genes
kshv.genes <- str_detect(rownames(normalized.counts), "^KSHV")

# Subset for KSHV genes
normalized.counts.kshv <- normalized.counts[kshv.genes, ]

export.kshv.gene.counts <- TRUE
if(export.kshv.gene.counts == TRUE){
  
  write.csv(normalized.counts.kshv, file = "Results/kshv.gene.counts.csv")
  
}

# Run PCA
kshv.pca <- pca(normalized.counts.kshv, 
                metadata = merged_pData, 
                removeVar = 0.1)

# Plot PCA

kshv.pca.plot.by.sample <- biplot(kshv.pca, 
       colby = contrast, 
       legendPosition = 'top', 
       legendLabSize = 16, 
       legendIconSize = 8.0)

screeplot <- screeplot(kshv.pca)

# Visualize clustering of genes

kshv_annotation <- as.data.frame(read_excel("kshv_genes_annotation.xlsx"))

rownames(kshv_annotation) <- kshv_annotation$gene

# Transpose gene counts
normalized.counts.kshv.transpose <- as.data.frame(t(normalized.counts.kshv))

# Ensure the rows and columns of the metadata and counts are aligned
kshv_annotation <- kshv_annotation[order(rownames(kshv_annotation)), ]
normalized.counts.kshv.transpose <- normalized.counts.kshv.transpose[ order(colnames(normalized.counts.kshv.transpose))]

# Create PCA for genes 

kshv.pca.by.gene <- pca(normalized.counts.kshv.transpose, 
                        metadata = kshv_annotation, 
                        removeVar = 0.1)


kshv.pca.plot.by.gene <- biplot(kshv.pca.by.gene, 
       colby = "type", 
       legendPosition = 'top', 
       legendLabSize = 16, 
       legendIconSize = 8.0)

export.PCA <- TRUE
if(export.PCA == TRUE){
  
  ggsave(paste0("Results/QC/kshv_pca_plot_by_sample_", contrast, ".png"),
         plot = kshv.pca.plot.by.sample, 
         width = 8, 
         height = 8, 
         units = "in")
  
  ggsave(paste0("Results/QC/kshv_pca_plot_by_gene.png"),
         plot = kshv.pca.plot.by.gene, 
         width = 8, 
         height = 8, 
         units = "in")
  
}

```
